---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then
     regenerate with devtools::build_readme(). Do not use knitr::knit(): it
     processes the code but leaves this YAML header in the output as literal
     text, which GitHub and pkgdown both render verbatim. -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/statmodels7/optimizers7/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/statmodels7/optimizers7/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/statmodels7/optimizers7/graph/badge.svg)](https://app.codecov.io/gh/statmodels7/optimizers7)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(optimizers7)
set.seed(1)
```

# optimizers7

Almost every R package that fits a model carries its own optimiser, written
inside the function that needs it: a loop, a `while (!converged)`, a tolerance
compared against whatever quantity the author had to hand. Nothing outside can
reuse it, extend it, or ask it for anything its author did not happen to need —
and the stopping rule, which decides what the whole thing means by *finished*,
is usually a number buried three levels down.

`{optimizers7}` writes them once, as objects. An optimiser carries its algorithm
and every setting that algorithm obeys; a stopping rule is a separate object you
can replace, combine, or write yourself; and the objective may be an R function,
a sum over observations, or compiled C++ reached through a pointer, with the
algorithms written once against all three.

It is the optimisation layer of [statmodels7](https://statmodels7.github.io),
alongside [linkfunctions7](https://statmodels7.github.io/linkfunctions7) and
[distributions7](https://statmodels7.github.io/distributions7).

## Installation

```r
# install.packages("pak")
pak::pak("statmodels7/optimizers7")
```

## The shape of it

Everything minimises, always, through one generic:

```{r}
f  <- function(p) 100 * (p[2] - p[1]^2)^2 + (1 - p[1])^2
gr <- function(p) c(-400 * p[1] * (p[2] - p[1]^2) - 2 * (1 - p[1]),
                    200 * (p[2] - p[1]^2))

minimize(bfgs(), f, par = c(-1.2, 1), gr = gr)
```

Swapping the algorithm changes one word, because every optimiser carries the
same settings:

```{r}
minimize(lbfgs(), f, c(-1.2, 1), gr = gr)@par
minimize(newton(), f, c(-1.2, 1), gr = gr)@par
minimize(nelder_mead(), f, c(-1.2, 1))@par
```

## The stopping rule is an object

This is the part that is not a convenience. A rule buried inside an algorithm
fixes, at the moment the package is written, what everyone downstream is allowed
to mean by convergence. Here it is a value:

```{r}
crit_any(crit_grad(1e-10), crit_rel_obj(1e-14))
```

Rules compose, and a rule the authors never thought of is a first-class citizen —
one class, one method:

```{r}
CritTiny <- S7::new_class("CritTiny", parent = criterion,
                          properties = list(tol = S7::class_numeric))
S7::method(crit_met, CritTiny) <- function(criterion, state)
  state$f_new < criterion@tol

minimize(bfgs(criterion = CritTiny(label = "f < 1e-8", tol = 1e-8)),
         f, c(-1.2, 1), gr = gr)@criterion_met
```

A rule an optimiser cannot evaluate is **refused**, by name, rather than
accepted and left never to fire:

```{r, error = TRUE}
minimize(nelder_mead(criterion = crit_grad()), f, c(-1.2, 1))
```

## Bounds are removed, not enforced

Each bounded coordinate is reparametrised onto the whole real line, so every
point proposed is admissible by construction — no rejection step, no boundary for
a line search to trip over, and any method gets bounds without knowing they
exist:

```{r}
y <- rnorm(200, mean = 0, sd = 2)
nll <- function(p) length(y) * log(p) + sum(y^2) / (2 * p^2)

minimize(bfgs(), nll, par = 1, bounds = list(c(0, Inf)))@par
sqrt(mean(y^2))
```

## When the derivative does not exist

A descent method walks to the minimum of a sum of absolute deviations and then
reports that it did not converge — correctly, because the subgradient it
evaluates there has norm 1 and never becomes small. `bundle()` keeps a
*collection* of subgradients and tests whether zero lies in their convex hull,
which is the statement that actually holds:

```{r}
z <- rnorm(101)
sad <- function(p) sum(abs(z - p))
sub <- function(p) -sum(sign(z - p))

c(bfgs   = minimize(bfgs(), sad, par = 0, gr = sub)@converged,
  bundle = minimize(bundle(), sad, par = 0, gr = sub)@converged)

c(bundle = minimize(bundle(), sad, par = 0, gr = sub)@par, median = median(z))
```

## What is here

| | |
|---|---|
| smooth, with derivatives | `newton()`, `bfgs()`, `lbfgs()`, `gradient_descent()` |
| stochastic, and for long parameter vectors | `adam()`, with minibatches over a `finite_sum()` |
| no derivative at all | `nelder_mead()`, `compass()` |
| non-smooth, with subgradients | `bundle()` |
| wrapping any of them | `multistart()` |
| line searches | `armijo()`, `wolfe()` |
| stopping rules | `crit_grad()`, `crit_abs_obj()`, `crit_rel_obj()`, `crit_abs_par()`, `crit_rel_par()`, `crit_stationary()`, `crit_never()`, and `crit_any()` / `crit_all()` to combine them |

Every method reports which safeguards fired, counts its own evaluations, and
sets `converged` only when the stopping rule was satisfied — never because the
iteration budget ran out.

## Checking an optimiser of your own

`check_optimizer()` verifies the contract — that `value` is the objective at
`par`, that a reported gradient is the gradient there, that bounds hold
*strictly*, that a run repeats — and then reports how the method did on the
standard test problems, separately, because those are different questions:

```{r}
check_optimizer(bfgs())
```

## Related

- [linkfunctions7](https://statmodels7.github.io/linkfunctions7) — link functions
  with exact derivatives to fourth order
- [distributions7](https://statmodels7.github.io/distributions7) — distributions
  carrying exact derivatives of the log-likelihood
- [the book](https://statmodels7.github.io/book/) — the mathematics, derived
