---
title: "Extending optimizers7"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending optimizers7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(optimizers7)
set.seed(1)
```

A toolkit that only worked for the algorithms it ships with would have solved
nothing. This is the other half: what you can add, and what the package
guarantees about the pieces you add.

There are three things you might want to replace — the stopping rule, the
objective, and the algorithm — and they are separate on purpose, so that
replacing one costs nothing in the other two.

## A stopping rule of your own

The alternative to an object here would be an argument taking a string and a
`switch` inside every algorithm. That arrangement fixes, at the moment the
package is written, what everyone downstream is allowed to mean by *finished* —
and convergence is not a detail, it is the definition of the answer.

A criterion is one class and one method. The method receives the whole state of
the iteration and returns a single logical:

```{r}
CritTiny <- S7::new_class("CritTiny", parent = criterion,
                          properties = list(tol = S7::class_numeric))

S7::method(crit_met, CritTiny) <- function(criterion, state) {
  state$f_new < criterion@tol
}

tiny <- CritTiny(label = "objective below 1e-8", tol = 1e-8)
```

It is then usable by every algorithm in the package, with no further ceremony:

```{r}
f  <- function(p) sum((p - c(1, 2))^2)
gr <- function(p) 2 * (p - c(1, 2))

minimize(bfgs(criterion = tiny), f, c(0, 0), gr = gr)@criterion_met
```

`state` carries `iter`, `f_new`, `f_old`, `x_new`, `x_old`, `gradient` and
`stationarity`. The last two are not always there, and that matters.

### Saying what your rule needs

A rule that reads the gradient, handed to a method that computes none, would sit
testing `NULL` at every iteration and quietly never fire; the run would end on
its iteration budget and report a reason nowhere near the truth. So a rule
declares what it reads, and an optimiser that cannot supply it refuses the rule
by name:

```{r, error = TRUE}
minimize(nelder_mead(criterion = crit_grad()), f, c(0, 0))
```

If your rule reads something optional, say so:

```{r}
S7::method(crit_needs, CritTiny) <- function(criterion) character()
crit_needs(crit_grad())
crit_needs(crit_stationary())
```

Rules combine, and the combination is itself a rule, so combinations nest:

```{r}
crit_any(tiny, crit_grad(1e-10))
```

## The objective, in whichever shape you have it

`minimize()` dispatches on the optimiser, so an algorithm is written once.
Which of three shapes the objective arrived in is settled separately, by
`as_objective()`, which dispatches on the objective. Each generic dispatches
where there is real variation; dispatching `minimize()` on both would need one
method per algorithm per shape.

### An ordinary function

The common case, and the one that costs the most: every evaluation is a callback
into R.

```{r}
minimize(bfgs(), f, c(0, 0), gr = gr)@counts
```

With no gradient supplied, one is obtained by central differences — and the
result says so, so a run is never silently less exact than it appears:

```{r}
minimize(bfgs(), f, c(0, 0))@message
```

### An objective that resamples itself

Subsampling is a property of the objective, not of an algorithm, and the
sharpest way to say so is that it needs no support at all. An objective that
draws a fresh minibatch on every call is an ordinary R function:

```{r}
y <- rnorm(2000, mean = 3)
m <- 100

batch    <- function(par) { i <- sample.int(2000, m); sum((y[i] - par)^2) / 2 }
batch_gr <- function(par) { i <- sample.int(2000, m); -sum(y[i] - par) }

minimize(adam(alpha = 0.05, decay = 0.005, maxit = 3000),
         batch, par = 0, gr = batch_gr)@par
mean(y)
```

There was once a `finite_sum()` class here, by which an objective declared that
it was a sum over observations so that `adam()` could ask it for a subset. It is
gone, and its removal is worth a paragraph because it is the same judgement
repeated all through this package.

Supporting it meant a second kind of objective, a rule for which stopping rules
that objective permitted, a way of reporting which was in force, and a branch in
the compiled loop — all so that the optimiser could do something the caller could
do in one line, and do better, since the caller is the one who knows what an
observation is. **An optimiser that knows about observations is an optimiser
that has stopped being a general one.** Draw the batch where the knowledge is.

One thing to get right when you do. Resample *inside* the objective, not around
the run: calling `minimize(adam(maxit = 1), ...)` in a loop resets \eqn{m} and
\eqn{v} to zero and the bias correction to \eqn{t = 1} at every call, so each one
takes a first step of length \eqn{\alpha} and the accumulated moments — which are
the whole of the method — are thrown away each time.

### Compiled C++

The reason the algorithms are written in C++ at all. Two external pointers, and
the iteration never returns to R:

```r
Rcpp::cppFunction(depends = "RcppArmadillo", '
  Rcpp::XPtr<double(*)(const arma::vec&)> make_fn() {
    return Rcpp::XPtr<double(*)(const arma::vec&)>(
      new double(*)(const arma::vec&)(&my_objective));
  }')

minimize(bfgs(), cpp_objective(fn = make_fn(), gr = make_gr()), par = c(0, 0))
```

The pointers are **tagged** rather than passed bare, and that is a safety
measure rather than a formality: a bare external pointer may be anything at all,
and calling through the wrong one is not an R error but a crash of the process.

```{r, error = TRUE}
cpp_objective(function(p) sum(p^2))
```

## An algorithm of your own

An optimiser is a subclass of `optimizer` carrying its own settings, plus a
method on `minimize()`. Here is one — Barzilai–Borwein, a first-order method
whose step length comes from the previous secant pair rather than from a line
search:

```{r}
BarzilaiBorwein <- S7::new_class("BarzilaiBorwein", parent = optimizer,
                                 properties = list(step = S7::class_numeric))

barzilai <- function(criterion = crit_grad(1e-8), step = 1e-3, maxit = 1000,
                     max_eval = 10000, verbose = FALSE, refresh = 50,
                     keep_trace = FALSE) {
  BarzilaiBorwein(name = "barzilai-borwein", criterion = criterion,
                  maxit = maxit, max_eval = max_eval, verbose = verbose,
                  refresh = refresh, keep_trace = keep_trace, step = step)
}

bb_run <- function(optimizer, fn, par, gr) {
  x <- as.numeric(par); g <- gr(x); a <- optimizer@step
  nf <- 0L; ng <- 1L; converged <- FALSE; it <- 0L

  for (it in seq_len(optimizer@maxit)) {
    x_new <- x - a * g
    g_new <- gr(x_new); ng <- ng + 1L
    s <- x_new - x; yv <- g_new - g
    denom <- sum(s * yv)
    # The safeguard, and it is not optional: a non-positive or vanishing
    # denominator means the secant pair carries no curvature, and dividing by
    # it produces a step of any size at all.
    a <- if (is.finite(denom) && denom > 0) sum(s * s) / denom else optimizer@step
    x <- x_new; g <- g_new

    state <- list(iter = it, f_new = fn(x), f_old = NULL, x_new = x,
                  x_old = x_new - s, gradient = g)
    nf <- nf + 1L
    if (crit_met(optimizer@criterion, state)) { converged <- TRUE; break }
  }

  optimizer_result(
    par = x, value = fn(x), gradient = g,
    counts = c(f = nf + 1L, g = ng, h = 0L), iterations = it,
    converged = converged,
    criterion_met = if (converged) optimizer@criterion@label
                    else "iteration budget reached",
    message = "", trace = NULL, optimizer = optimizer, elapsed = NA_real_
  )
}

S7::method(minimize, BarzilaiBorwein) <-
  function(optimizer, fn, par, gr = NULL, he = NULL, bounds = NULL, ...) {
    if (is.null(gr)) stop("barzilai() needs a gradient.", call. = FALSE)
    bb_run(optimizer, fn, par, gr)
  }
```

Note the thing the method is responsible for and that nothing can do for it:
`converged` is set from whether the rule fired and **never** from the run having
ended.

It works, and on a quadratic it is startlingly good — two iterations, because
the Barzilai–Borwein step length is exactly the inverse curvature when the
curvature is constant:

```{r}
r <- minimize(barzilai(), f, c(0, 0), gr = gr)
c(par = r@par, iterations = r@iterations, converged = r@converged)
```

## Checking it

`check_optimizer()` exists for exactly this moment. It separates two questions
that are easy to confuse: whether the optimiser keeps its **contract**, and how
**powerful** it is. Those are different, and conflating them would make the
function useless — gradient descent does not solve Rosenbrock in five hundred
iterations and is not broken, it is slow.

So the numbered checks are statements every optimiser must satisfy however weak
it is, and the battery underneath is a table of gaps, information rather than
verdict:

```{r}
res <- check_optimizer(barzilai(step = 1e-3))
```

Read the battery first: this method is *strong*. It solves every smooth problem
in the list to machine precision, and is beaten only by the two that are
multimodal and the one with a kink — which is correct behaviour, not failure.

And read the two failed checks second, because they are the more useful half.
Nothing above is wrong with the *algorithm*. What is wrong is that the method
accepted two arguments and did nothing with them.

**It ignored `bounds`.** The signature has to include it — S7 requires a
method's formals to contain every named argument of the generic — and the
easiest thing to do with an argument you did not plan for is to let it sit
there. A caller then passes a box, gets no error, and receives an answer outside
it. Removing the box is four lines, because
`bounded_transform()` does the work:

```{r}
suppressMessages(
S7::method(minimize, BarzilaiBorwein) <-
  function(optimizer, fn, par, gr = NULL, he = NULL, bounds = NULL, ...) {
    if (is.null(gr)) stop("barzilai() needs a gradient.", call. = FALSE)
    check_criterion(optimizer)                                     # (2)

    if (is.null(bounds)) return(bb_run(optimizer, fn, par, gr))    # (1)

    # Optimise in eta, where there is no box, and map back to report. The
    # Jacobian is diagonal, so the chain rule is one multiplication.
    hh <- function(e) vapply(seq_along(e),
            function(j) bounded_transform(bounds[[j]], e[j])$h, 0)
    dd <- function(e) vapply(seq_along(e),
            function(j) bounded_transform(bounds[[j]], e[j])$d1, 0)
    eta0 <- vapply(seq_along(par),
                   function(j) bounded_forward(bounds[[j]], par[j]), 0)

    res <- bb_run(optimizer,
                  function(e) fn(hh(e)),
                  eta0,
                  function(e) gr(hh(e)) * dd(e))

    res@gradient <- res@gradient / dd(res@par)   # both on the user's scale
    res@par <- hh(res@par)
    res
  }
)
```

**It never validated the stopping rule.** `check_criterion()` — marked (2) — is
the one line that refuses a rule the method cannot evaluate, instead of
accepting one that would test `NULL` for ever. What a method *can* supply is
declared by `optimizer_provides()`, whose default claims a gradient and an
objective. This one has both, so the default is honest; a derivative-free method
of your own would need one line:

```r
S7::method(optimizer_provides, BarzilaiBorwein) <- function(optimizer)
  c("objective", "stationarity")
```

With both promises kept, the contract holds:

```{r}
all(check_optimizer(barzilai(step = 1e-3), problems = test_problems("sphere"),
                    verbose = FALSE)$checks)
```

That is the whole point of the numbered checks. The algorithm was right from the
first line; what was missing was two promises the package makes on every
optimiser's behalf, and neither omission would have shown itself in any run that
happened to succeed.

## The test problems are yours too

`test_problems()` is exported because a battery is useful to anyone writing an
optimiser, not only to this package:

```{r}
p <- test_problems("beale")[[1]]
str(p, max.level = 1)

minimize(bfgs(), p$fn, p$par, gr = p$gr)@par
p$solution
```

Two of them are marked `multimodal`, where a local method reaching a different
minimum is behaving correctly rather than failing, and one is marked
non-`smooth`, where a method that assumes a derivative exists will walk to the
answer and then be unable to certify it:

```{r}
sapply(test_problems(), function(p) c(multimodal = p$multimodal,
                                      smooth = p$smooth))
```
