---
title: "Extending optimizers7"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending optimizers7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(optimizers7)
set.seed(1)
```

A toolkit that only worked for the algorithms it ships with would have solved
nothing. This is the other half: what you can add, and what the package
guarantees about the pieces you add.

There are three things you might want to replace — the stopping rule, the
objective, and the algorithm — and they are separate on purpose, so that
replacing one costs nothing in the other two.

## A stopping rule of your own

The alternative to an object here would be an argument taking a string and a
`switch` inside every algorithm. That arrangement fixes, at the moment the
package is written, what everyone downstream is allowed to mean by *finished* —
and convergence is not a detail, it is the definition of the answer.

A criterion is one class and one method. The method receives the whole state of
the iteration and returns a single logical:

```{r}
CritTiny <- S7::new_class("CritTiny", parent = criterion,
                          properties = list(tol = S7::class_numeric))

S7::method(crit_met, CritTiny) <- function(criterion, state) {
  state$f_new < criterion@tol
}

tiny <- CritTiny(label = "objective below 1e-8", tol = 1e-8)
```

It is then usable by every algorithm in the package, with no further ceremony:

```{r}
f  <- function(p) sum((p - c(1, 2))^2)
gr <- function(p) 2 * (p - c(1, 2))

minimize(bfgs(criterion = tiny), f, c(0, 0), gr = gr)@criterion_met
```

`state` carries `iter`, `f_new`, `f_old`, `x_new`, `x_old`, `gradient` and
`stationarity`. The last two are not always there, and that matters.

### Saying what your rule needs

A rule that reads the gradient, handed to a method that computes none, would sit
testing `NULL` at every iteration and quietly never fire; the run would end on
its iteration budget and report a reason nowhere near the truth. So a rule
declares what it reads, and an optimiser that cannot supply it refuses the rule
by name:

```{r, error = TRUE}
minimize(nelder_mead(criterion = crit_grad()), f, c(0, 0))
```

If your rule reads something optional, say so:

```{r}
S7::method(crit_needs, CritTiny) <- function(criterion) character()
crit_needs(crit_grad())
crit_needs(crit_stationary())
```

Rules combine, and the combination is itself a rule, so combinations nest:

```{r}
crit_any(tiny, crit_grad(1e-10))
```

## The objective

`minimize()` dispatches on the optimiser, so an algorithm is written once. The
objective is normalised separately, by `as_objective()`, which dispatches on the
objective. Each generic dispatches where there is real variation; dispatching
`minimize()` on both would need one method per algorithm per shape.

### An ordinary function

The common case, and the one that costs the most: every evaluation is a callback
into R.

```{r}
minimize(bfgs(), f, c(0, 0), gr = gr)@counts
```

With no gradient supplied, one is obtained by central differences — and the
result says so, so a run is never silently less exact than it appears:

```{r}
minimize(bfgs(), f, c(0, 0))@message
```

### An objective that resamples itself

Subsampling is a property of the objective, not of an algorithm, and the
sharpest way to say so is that it needs no support at all. An objective that
draws a fresh minibatch on every call is an ordinary R function:

```{r}
y <- rnorm(2000, mean = 3)
m <- 100

batch    <- function(par) { i <- sample.int(2000, m); sum((y[i] - par)^2) / 2 }
batch_gr <- function(par) { i <- sample.int(2000, m); -sum(y[i] - par) }

minimize(adam(alpha = 0.05, decay = 0.005, maxit = 3000),
         batch, par = 0, gr = batch_gr)@par
mean(y)
```

There was once a `finite_sum()` class here, by which an objective declared that
it was a sum over observations so that `adam()` could ask it for a subset. It is
gone, and its removal is worth a paragraph because it is the same judgement
repeated all through this package.

Supporting it meant a second kind of objective, a rule for which stopping rules
that objective permitted, a way of reporting which was in force, and a branch in
the compiled loop — all so that the optimiser could do something the caller could
do in one line, and do better, since the caller is the one who knows what an
observation is. **An optimiser that knows about observations is an optimiser
that has stopped being a general one.** Draw the batch where the knowledge is.

One thing to get right when you do. Resample *inside* the objective, not around
the run: calling `minimize(adam(maxit = 1), ...)` in a loop resets \eqn{m} and
\eqn{v} to zero and the bias correction to \eqn{t = 1} at every call, so each one
takes a first step of length \eqn{\alpha} and the accumulated moments — which are
the whole of the method — are thrown away each time.

### Your own kind of objective

There is one shipped shape, and `as_objective()` is a generic so that there can
be others. A caller holding a fitted-model object, say, registers one method and
every algorithm in the package accepts it:

```r
S7::method(as_objective, MyModel) <- function(fn, gr = NULL, he = NULL, ...) {
  list(kind = "r",
       fn = function(par) -loglik(fn, par),
       gr = function(par) -score(fn, par),
       he = NULL,
       has_gradient = TRUE, has_hessian = FALSE)
}
```

There was a second shipped shape, `cpp_objective()`, taking a pair of external
pointers so the iteration never returned to R. It is gone, and the reasoning is
worth recording because it is easy to assume the opposite.

It could not carry data. The pointer type was `double(*)(const arma::vec&)` — a
bare function pointer with no closure and no user-data argument — so any real
statistical objective had to keep its `y` and its `X` in C++ globals. That is
not reentrant, it means one model at a time, and it rules out running several
starts in one process.

And it did not pay. Measured on a Gamma regression against the same objective
written in vectorised R, the compiled version was *slower* below about twenty
thousand observations and never better than about twice as fast above it. The
reason is structural rather than incidental: assembling a gradient or a Hessian
from model terms is `crossprod(Z, s)` and `crossprod(Z, W * Z)`, and R's matrix
arithmetic and Armadillo's call the same BLAS. On the operation that dominates a
real fit the two languages are the same code.

Compiled kernels earn their keep where the arithmetic is genuinely elementwise
and irregular — which is exactly where \pkg{distributions7} puts them, in the
fourth-order derivative expressions of each family. Not here.

## An algorithm of your own

An optimiser is a subclass of `optimizer` carrying its own settings, plus a
method on `minimize()`. Here is one the package does not ship — the heavy-ball
method, gradient descent with momentum, which adds a fraction of the previous
*displacement* to each step:

$$x_{k+1} = x_k - lpha g_k + eta (x_k - x_{k-1}).$$

The momentum term is what carries the iterate along a valley floor instead of
across it, and it is the ancestor of the moment estimate in `adam()`.

```{r}
HeavyBall <- S7::new_class("HeavyBall", parent = optimizer,
                           properties = list(alpha = S7::class_numeric,
                                             beta  = S7::class_numeric))

heavy_ball <- function(criterion = crit_grad(1e-8), alpha = 1e-3, beta = 0.9,
                       maxit = 2000, max_eval = 20000, verbose = FALSE,
                       refresh = 50, keep_trace = FALSE) {
  HeavyBall(name = "heavy ball", criterion = criterion,
            maxit = maxit, max_eval = max_eval, verbose = verbose,
            refresh = refresh, keep_trace = keep_trace,
            alpha = alpha, beta = beta)
}

hb_run <- function(o, fn, par, gr) {
  x <- as.numeric(par); x_prev <- x
  nf <- 0L; ng <- 0L; converged <- FALSE; it <- 0L

  for (it in seq_len(o@maxit)) {
    g <- gr(x); ng <- ng + 1L
    x_new <- x - o@alpha * g + o@beta * (x - x_prev)
    # The safeguard, and it is not optional: momentum can carry the iterate
    # somewhere the objective is undefined, and one NaN contaminates every
    # iterate after it.
    if (!all(is.finite(x_new))) break
    x_prev <- x; x <- x_new

    state <- list(iter = it, f_new = fn(x), f_old = fn(x_prev), x_new = x,
                  x_old = x_prev, gradient = gr(x))
    nf <- nf + 2L; ng <- ng + 1L
    if (crit_met(o@criterion, state)) { converged <- TRUE; break }
  }

  optimizer_result(
    par = x, value = fn(x), gradient = gr(x),
    counts = c(f = nf + 1L, g = ng + 1L, h = 0L), iterations = it,
    converged = converged,
    criterion_met = if (converged) o@criterion@label
                    else "iteration budget reached",
    message = "", trace = NULL, optimizer = o, elapsed = NA_real_
  )
}

S7::method(minimize, HeavyBall) <-
  function(optimizer, fn, par, gr = NULL, he = NULL, bounds = NULL, ...) {
    if (is.null(gr)) stop("heavy_ball() needs a gradient.", call. = FALSE)
    hb_run(optimizer, fn, par, gr)
  }
```

Note the thing the method is responsible for and that nothing can do for it:
`converged` is set from whether the rule fired and **never** from the run having
ended.

It works:

```{r}
r <- minimize(heavy_ball(alpha = 0.05), f, c(0, 0), gr = gr)
c(par = r@par, iterations = r@iterations, converged = r@converged)
```

## Checking it

`check_optimizer()` exists for exactly this moment. It separates two questions
that are easy to confuse: whether the optimiser keeps its **contract**, and how
**powerful** it is. Those are different, and conflating them would make the
function useless — gradient descent does not solve Rosenbrock in five hundred
iterations and is not broken, it is slow.

So the numbered checks are statements every optimiser must satisfy however weak
it is, and the battery underneath is a table of gaps, information rather than
verdict:

```{r}
res <- check_optimizer(heavy_ball(alpha = 0.01))
```

Read the battery first, and read it as information rather than as a verdict.
A fixed step length is a real limitation: on the problems whose curvature varies
by orders of magnitude this method crawls, and that is a fair description of
heavy ball rather than a defect in it. `bb()` — which the package does ship —
is the same shape of method with the step length estimated from the last secant
pair instead of fixed, and it solves those same problems to machine precision.

And read the two failed checks second, because they are the more useful half.
Nothing above is wrong with the *algorithm*. What is wrong is that the method
accepted two arguments and did nothing with them.

**It ignored `bounds`.** The signature has to include it — S7 requires a
method's formals to contain every named argument of the generic — and the
easiest thing to do with an argument you did not plan for is to let it sit
there. A caller then passes a box, gets no error, and receives an answer outside
it. Removing the box is four lines, because
`bounded_transform()` does the work:

```{r}
suppressMessages(
S7::method(minimize, HeavyBall) <-
  function(optimizer, fn, par, gr = NULL, he = NULL, bounds = NULL, ...) {
    if (is.null(gr)) stop("heavy_ball() needs a gradient.", call. = FALSE)
    check_criterion(optimizer)                                     # (2)

    if (is.null(bounds)) return(hb_run(optimizer, fn, par, gr))    # (1)

    # Optimise in eta, where there is no box, and map back to report. The
    # Jacobian is diagonal, so the chain rule is one multiplication.
    hh <- function(e) vapply(seq_along(e),
            function(j) bounded_transform(bounds[[j]], e[j])$h, 0)
    dd <- function(e) vapply(seq_along(e),
            function(j) bounded_transform(bounds[[j]], e[j])$d1, 0)
    eta0 <- vapply(seq_along(par),
                   function(j) bounded_forward(bounds[[j]], par[j]), 0)

    res <- hb_run(optimizer,
                  function(e) fn(hh(e)),
                  eta0,
                  function(e) gr(hh(e)) * dd(e))

    res@gradient <- res@gradient / dd(res@par)   # both on the user's scale
    res@par <- hh(res@par)
    res
  }
)
```

**It never validated the stopping rule.** `check_criterion()` — marked (2) — is
the one line that refuses a rule the method cannot evaluate, instead of
accepting one that would test `NULL` for ever. What a method *can* supply is
declared by `optimizer_provides()`, whose default claims a gradient and an
objective. This one has both, so the default is honest; a derivative-free method
of your own would need one line:

```r
S7::method(optimizer_provides, HeavyBall) <- function(optimizer)
  c("objective", "stationarity")
```

With both promises kept, the contract holds:

```{r}
all(check_optimizer(heavy_ball(alpha = 0.01),
                    problems = test_problems("sphere"),
                    verbose = FALSE)$checks)
```

That is the whole point of the numbered checks. The algorithm was right from the
first line; what was missing was two promises the package makes on every
optimiser's behalf, and neither omission would have shown itself in any run that
happened to succeed.

## The test problems are yours too

`test_problems()` is exported because a battery is useful to anyone writing an
optimiser, not only to this package:

```{r}
p <- test_problems("beale")[[1]]
str(p, max.level = 1)

minimize(bfgs(), p$fn, p$par, gr = p$gr)@par
p$solution
```

Two of them are marked `multimodal`, where a local method reaching a different
minimum is behaving correctly rather than failing, and one is marked
non-`smooth`, where a method that assumes a derivative exists will walk to the
answer and then be unable to certify it:

```{r}
sapply(test_problems(), function(p) c(multimodal = p$multimodal,
                                      smooth = p$smooth))
```
